---
title: "Relational algebra as a tool to help on your SQL and dplyr journey"
author: "L T Droy"
date: "30/05/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Relational Algebra

Despite routinely doing data manipulation tasks in R, I only recently discovered *relational algebra* as a way of thinking in a *formal* way about these tasks. I came across relational algebra as part of my ongoing CPD journey, which has recently focused on learning PostgreSQL.

Relational algebra provides a formal logic and notation for thinking about operations which manipulate and join (etc.) tables of data. Building or using an algebra of table manipulations might sound like a dry academic exercise. However, I've found that learning some of the basic of relational algebra is very practically useful:

1. It helps me to think more clearly about data manipulation operations.
2. It helps me to understand when different sets of data manipulation code are logically equivilent (and so will produce the same result).
3. When using R, it helps me to write for efficient code. 
4 (bonus). It provides the basis for a educational coding challenge!

## A brief overview of relational algebra


## 4 (Bonus) Educational coding challenge

```{r}

library(purrr)
library(digest)

R1 <- data.frame(
  a=c(1,2,3,4,5),
  b=c("a","b","c","d","e"),
  c=c(TRUE,FALSE,TRUE,FALSE,FALSE),
  stringsAsFactors = FALSE
)

R2 <- data.frame(
  c=c(1,2,3,4),
  d=c("a","b","c","d"),
  stringsAsFactors = FALSE
)

project <- function(R, names){
  return(R[,names])
}

select <- function(R, cond_func){
  return(R[cond_func(R),])
}

cond_func_eq <- function(attr1, attr2){
  return(
    function(R){
      R[[attr1]] == R[[attr2]]
    }
  )
}

rename <- function(R, new_names){
  names(R) <- new_names
}

cross_product <- function(R1, R2){
  
  R1_names <- names(R1)
  R2_names <- names(R2)
  
  new_names <- c(names(R1), names(R2))
  
  is_duplicated <- function(nm, nms){
    return(sum(nms==nm) > 1)
  }
  
  dup_r1_index <- map_lgl(R1_names, is_duplicated, new_names)
  dup_r2_index <- map_lgl(R2_names, is_duplicated, new_names)
  
  names(R1)[dup_r1_index] <- paste0("L_", names(R1)[dup_r1_index])

  names(R2)[dup_r2_index] <- paste0("R_", names(R2)[dup_r2_index])

  indexes <- expand.grid(1:nrow(R2), 1:nrow(R1))
  
  R1 <- R1[indexes[[2]], ]
  R2 <- R2[indexes[[1]], ]

  return(cbind(R1, R2))
  
}

inner_join <- function(R1, R2, cond_func){
  
  R3 <- cross_product(R1, R2)
  
  return(R3[cond_func(R3),])
  
}

ra_union <- function(R1, R2){
  check_set_compatibility(R1, R2)
  R3 <- rbind(R1, R2)
  names(R3) <- names(R1)
  return(R3)
}

ra_diff <- function(R1, R2){
  check_set_compatibility(R1, R2)
  R1_hashes <- apply(R1, 1, digest)
  R2_hashes <- apply(R2, 1, digest)
  R1_hash_diff <- setdiff(R1_hashes, R2_hashes)
  return(R1[R1_hashes %in% R1_hash_diff, ])
}

ra_group <- function(R1, group_columns, cell_columns, func){
  
  #TODO add grouping method
  
}

check_set_compatibility <- function(R1, R2){
  
  R1_col_modes <- as.character(apply(R1, 2, mode))
  R2_col_modes <- as.character(apply(R2, 2, mode))
  
  if(!( ncol(R1) == ncol(R2) ) | !(identical(R1_col_modes, R2_col_modes)) ){
    stop("Data frames are not compatible.")
  }
}







```



